Ultrafast manipulation of Heisenberg exchange and Dzyaloshinskii Moriya interactions in antiferromagnetic insulators

### 2019-05-02 Presentation/presentation.tex
presentation updates

%\author{Juan Manuel Losada Sosnovsky} % Your name
%\autoref{asdf}{asdf}
%\author[author1]{author1\inst{1}\\[1ex]  {\small supervisor1\inst{1,2}}}
\author[Juan Manuel Losada Sosnovsky]{Juan Manuel Losada Sosnovsky\\[10mm]{\small Supervisor: Alireza Qaiumzadeh}}
\institute[NTNU] % Your institution as it will appear on the bottom of every slide, may be shorthand to save space
{
QuSpin (NTNU)
}
\date{\today} % Date, can be changed to a custom date
\newcommand{\bs}[1] {\boldsymbol{#1}}
\begin{document}

@@

\begin{frame}
\frametitle{Manipulation of the spin state}
\begin{block}{Heisenberg model}
\begin{equation} 
\hat{H} = \sum_{\langle i,j \rangle} J \boldsymbol{S}_i \boldsymbol{S}_j \nonumber
\end{equation}
In antiferromagnets $J > 0$ ($J < 0$ for ferromagnets). Mentink, J. H. et al. showed that it is possible to manipulate the exchange constant by applying a laser field. $J = J_0 + \Delta J$.
\end{block}
\end{frame}

@@

\begin{frame}
\frametitle{Manipulation of the spin state}
\begin{block}{Heisenberg model with DMI}
\begin{equation} 
\hat{H} = \sum_{\langle i,j \rangle} J \boldsymbol{S}_i \boldsymbol{S}_j + \sum_{\langle i,j \rangle} \vec{D}_{ij}\boldsymbol{S}_i \times \boldsymbol{S}_j\nonumber
\end{equation}

Under some circumstances this Hamiltonian describes a canted antiferromagnet.

@@

\begin{frame}
\frametitle{Manipulation of the spin state}
If we can manipulate $J$ and $D$ we can control the canting angle.
\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/canted_afm.png}
\caption{Source: Kimmel presentation}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Manipulation of the spin state}
We need $\frac{J}{D} \neq \frac{J+\Delta J}{D+\Delta D}$.
\\[10mm]
However, $\frac{J}{D} = \frac{J+\Delta J}{D+\Delta D}$.
\end{frame}

\begin{frame}
\frametitle{Manipulation of the spin state}
\begin{block}{Kane-Mele-Hubbard model}
\begin{equation}
\hat{H}_{\text{KMH}} = -t_1\sum_{\langle i j \rangle \sigma} \hat{c}^{\dagger}_{i\sigma}\hat{c}_{j\sigma} + i\Delta \sum_{\langle \langle i j \rangle \rangle \sigma \sigma'} \hat{c}^{\dagger}_{i\sigma} \nu_{ij} \sigma^z_{\sigma \sigma'} \hat{c}_{j\sigma'} + \text{U}\hat{D}\nonumber
\end{equation}

It leads to a spin Hamiltonian:

\begin{equation}
\hat{H}_{\text{KMH}}^{\text{eff}} = \sum_{\langle i,j \rangle} J_{ij} \bs{S}_i \bs{S}_j + \sum_{\langle \langle i,j \rangle \rangle} \bs{S}_i \bs{\Gamma}_{ij} \bs{S}_j \nonumber
\end{equation}

@@

\begin{frame}
\frametitle{Manipulation of the spin state}
\begin{block}{Modified Kane-Mele-Hubbard model}
\begin{equation}
\hat{H} = - t_1\sum_{\langle i,j \rangle, \sigma} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} + 
	\sum_{\langle \langle i,j \rangle \rangle, \sigma}(t_2 + i\Delta\nu_{ij}\sigma^z_{\sigma, \sigma})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} + 
	\text{U}\hat{D}\nonumber
\end{equation}

It leads to a spin Hamiltonian:

\begin{equation}
\hat{H}_{\text{eff}}(t) = \sum_{\langle i,j \rangle} J_{1,ij}\bs{S}_i\bs{S}_j + \sum_{\langle \langle i,j \rangle \rangle} \left\{ J_{2,ij}\bs{S}_i\bs{S}_j + \bs{D}_{2,ij} \bs{S}_i \times \bs{S}_j + \bs{S}_i \bs{\Gamma}_{ij} \bs{S}_j \right\} \nonumber
\end{equation}

What we wanted, in this case if we apply light $\frac{J}{D} \neq \frac{J+\Delta J}{D + \Delta D}$.

>>>>

%\author{Juan Manuel Losada Sosnovsky} % Your name
%\autoref{asdf}{asdf}
%\author[author1]{author1\inst{1}\\[1ex]  {\small supervisor1\inst{1,2}}}
\author[Juan Manuel Losada Sosnovsky]{Juan Manuel Losada Sosnovsky\\[5mm]{\small Supervisors: Arne Brataas and Alireza Qaiumzadeh}}
\institute[NTNU] % Your institution as it will appear on the bottom of every slide, may be shorthand to save space
{
QuSpin (NTNU)
}
\date{\today} % Date, can be changed to a custom date
\newcommand{\bs}[1] {\boldsymbol{#1}}
\begin{document}

@@

\begin{frame}
\frametitle{Spin models}
\begin{block}{Heisenberg model}
\begin{equation} 
\hat{H} = \sum_{\langle i,j \rangle} J \bs{S}_i \bs{S}_j \nonumber
\end{equation}
In antiferromagnets $J > 0$ ($J < 0$ for ferromagnets). Mentink, J. H. et al. showed that it is possible to manipulate the exchange constant by applying a laser field. $J = J_0 + \Delta J$.
\end{block}
\end{frame}

@@

\begin{frame}
\frametitle{Spin models}
The Dzyaloshinskii-Moriya interaction (DMI):
\begin{equation*}
\bs{D}_{ij}\bs{S}_i\times\bs{S}_j
\end{equation*}
\begin{itemize}
\item Arises from the spin orbit coupling.
\item It favours canted order, i.e. weak ferromagnetism in AFM.
\item Enables topological objects such as chiral skyrmions and chiral domain walls.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Spin models}
\begin{block}{Heisenberg model with DMI}
\begin{equation} 
\hat{H} = \sum_{\langle i,j \rangle} J \bs{S}_i \bs{S}_j + \sum_{\langle i,j \rangle} \vec{D}_{ij}\bs{S}_i \times \bs{S}_j\nonumber
\end{equation}

The ratio between the exchange interaction and the DMI controls the tilt angle of the canted spins.

@@

\begin{frame}
\frametitle{Spin models}
If we can manipulate $J$ and $D$ we can control the canting angle.
\begin{figure}
\includegraphics[width=1\linewidth]{../Figures/canted_afm.png}
\caption{Source: Kimmel presentation}
\end{figure}
\end{frame}

%\begin{frame}
%\frametitle{Manipulation of the spin state}
%We need $\frac{J}{D} \neq \frac{J+\Delta J}{D+\Delta D}$.
%\\[10mm]
%However, $\frac{J}{D} = \frac{J+\Delta J}{D+\Delta D}$.
%\end{frame}

\begin{frame}
\frametitle{Model}
Strongly correlated electrons in a planar honeycomb lattice can be described by:
\begin{block}{Kane-Mele-Hubbard model}
\begin{equation}
\hat{H} = - t_1\sum_{\langle i,j \rangle, \sigma} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma} -
	\sum_{\langle \langle i,j \rangle \rangle, \sigma, \sigma'}(\delta_{\sigma,\sigma'}t_2 - i\Delta\nu_{ij}\sigma^z_{\sigma, \sigma'})\hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'} + \text{U}\hat{D}\nonumber
\end{equation}

@@

\begin{frame}
\frametitle{Model}
The effective spin Hamiltonian in the strongly correlated regime takes the form:
\begin{equation}
\hat{H}_{\text{eff}} = \sum_{\langle i,j \rangle} J_{1}\bs{S}_i\bs{S}_j + \sum_{\langle \langle i,j \rangle \rangle} \left\{ J_{2}\bs{S}_i\bs{S}_j + \bs{D}_{2,ij} \bs{S}_i \times \bs{S}_j + \bs{S}_i \bs{\Gamma} \bs{S}_j \right\} \nonumber
\end{equation}
With:
\begin{align*}
&J_{1(2)} = \frac{2t_{1(2)}^2}{U}, \\
&\bs{\Gamma} =\frac{2\Delta^2}{U} \text{diag}(-1,-1,1),\\
&\bs{D}_{ij} = \frac{4 t_2 \Delta}{U}\nu_{ij}  \hat{\mathrm{e}}_z.
\end{align*}

\begin{frame}
\frametitle{Method}
\begin{equation*}
\hat{H} = -\hat{T} + \text{U}\hat{D}
\end{equation*}
$\hat{T}$ is a hopping operator, and $\hat{D}$ is the doublon number operator.
\begin{align*}
\hat{D} &= \sum_{i=1}^M \hat{n}_{i\uparrow}\hat{n}_{i\downarrow} \\
\hat{T} &= \sum_{i,j, \sigma, \sigma'} t_{ij}^{\sigma \sigma'} \hat{c}_{i \sigma}^\dagger \hat{c}_{j \sigma'}
\end{align*}
$\text{U} \gg t_{ij}^{\sigma \sigma'}$, strongly correlated regime.
\end{frame}

\begin{frame}
\frametitle{Method}
To account for the laser illumination we use several techniques:
\begin{itemize}
\item Peierls substitution.
\item Apply time dependent unitary transformation.
\item Expand unitary transformation and transformed Hamiltonian.
\item Determine unitary transformation.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Peierls substitution}
In the presence of a vector potential $\bs{A}(t)$ the hopping amplitudes acquire a time dependent phase:
\begin{equation*}
\label{TimeDepHopping}
t_{ij}^{\sigma \sigma'}(t) = t_{ij}^{\sigma \sigma'} e^{ie\bs{R}_{ij}\bs{A}(t)}
\end{equation*}
The Hamiltonian is time dependent.
\end{frame}

\begin{frame}
\frametitle{Unitary transformation}
Apply $\hat{U}(t) = e^{-i\hat{S}(t)}$,

\begin{equation*}
\hat{H}'(t) = e^{i\hat{S}(t)} \hat{H}(t) e^{-i\hat{S}(t)} - e^{i\hat{S}(t)} id_t e^{-i\hat{S}(t)}
\end{equation*}
The goal is $\hat{H}'(t)$ to be block diagonal in $\hat{D}$ up to certain order in the hopping amplitudes.
\end{frame}

\begin{frame}
\frametitle{Expand transformed Hamiltonian}
Formally $\hat{T}(t) = \eta\hat{T}(t)$, where $\eta$ plays the role of a bookkeeping parameter in the perturbative expansion.
\begin{align*}
\hat{S}(t) &= \sum_\nu \eta^\nu \hat{S}^{(\nu)}(t) \\
\hat{H}'(t) &= \sum_\nu \eta^\nu \hat{H}'^{(\nu)}(t)
\end{align*}

The transformed Hamiltonian can be written:

\begin{equation*} 
\hat{H}'(t) = \sum_m \frac{1}{m!} \text{ad}_{i\hat{S}(t)}^m \left( \hat{H}(t) - \sum_n \frac{1}{(n+1)!}\text{ad}_{-i\hat{S}(t)}^n (d_t \hat{S}(t)) \right)
\end{equation*}
\end{frame}

\begin{frame}
\frametitle{Determine unitary transformation}
We decompose the hopping operator as $\hat{T}(t) = \sum_{d,m} \hat{T}_{d,m}e^{im\omega t}$, where $\hat{T}_{d,m}$ changes the doublon number by $d$.
\begin{align*}
\hat{H}'^{(0)}(t) &= \text{U}\hat{D} \\
\hat{H}'^{(1)}(t) &= -\sum_m \hat{T}_{0,m}(t)e^{im\omega t} \\
\hat{H}'^{(2)}(t) &= \frac{1}{2}\sum_{mn} \left( \frac{\left[\hat{T}_{1n}, \hat{T}_{-1(m-n)} \right]}{\text{U}+n\omega} - \frac{\left[\hat{T}_{-1n}, \hat{T}_{1(m-n)} \right]}{\text{U}-n\omega} \right) e^{im\omega t}
\end{align*}
\end{frame}

\begin{frame}
\frametitle{Projecting onto the low energy subspace}
In the strongly correlated limit we can assume that the Hamiltonian acts only within the $d=0$ subspace, i.e. $\hat{H}_{\text{eff}}=\langle \hat{P}_0\hat{H}'(t)\hat{P}_0 \rangle$. Obtaining:
\begin{equation*}
\hat{H}_{\text{eff}} = - \sum_{i,j, \sigma_1, \sigma_2, \sigma_3, \sigma_4}\hat{c}_{i \sigma_1}^\dagger \hat{c}_{j \sigma_2} \hat{c}_{j \sigma_3}^\dagger \hat{c}_{i \sigma_4} t_{ij}^{\sigma_1 \sigma_2} t_{ji}^{\sigma_3 \sigma_4} \left\{ \sum_{n} \frac{\mathcal{J}_{n}^2(\alpha_{ij})}{\text{U}+n\omega} \right\}
\end{equation*}
Where $\alpha_{ij} = e\bs{R}_{ij}\bs{A}$.
\end{frame}

\begin{frame}
\frametitle{Results}
If we apply this procedure to the Kane Mele Hubbard model, we get:
\begin{align*}
\tilde{H}_{\text{S}}(\omega) =& \sum_{\langle i,j \rangle} \tilde{J}_{1,ij} \bs{S}_i\bs{S}_j + \sum_{\langle \langle i,j \rangle \rangle} \left\{ \tilde{J}_{2,ij}\bs{S}_i\bs{S}_j + \bs{S}_i \tilde{\bs{\Gamma}}_{ij} \bs{S}_j + \tilde{\bs{D}}_{ij} \bs{S}_i \times \bs{S}_j \right\}
\end{align*}
with the following renormalized spin interactions:
\begin{align*}
&\tilde{J}_{1(2),ij} = 2t_{1(2)}^2\sum_n\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{U+n\omega}, \\
&\tilde{\bs{\Gamma}}_{ij} = 2\Delta^2 \text{diag}(-1,-1,1) \sum_n\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{U+n\omega},\\
&\tilde{\bs{D}}_{ij} = 4 t_2 \Delta \sum_n\frac{\mathcal{J}_{n}^2(\alpha_{ij})}{U+n\omega} \nu_{ij} \hat{\mathrm{e}}_z.
\end{align*}
\end{frame}
